Trail Saving on Backtrack

### 2020-05-09 paper.tex
reviewer 1 changes

\begin{document}
\maketitle
\begin{abstract}
    A CDCL \sat solver can backtrack a large distance when it learns a
    new clause, e.g, when the new learnt clause is a unit clause the
    solver has to backtrack to level zero. When the length of the
    backtrack is large, the solver can end up reproducing many of the
    same decisions and propagations when it redescends the search
    tree. Motivated by this potential redundancy, partial backtracking
    (chronological backtracking) has been proposed. This technique
    allows the solver to reduce the length of its backtrack, e.g.,
    allowing the solver to backtrack to the previous level after a
    conflict is found. However, this technique has its shortcomings
    and only tends to work when limits are placed on its
    application. In this paper we present a new trail saving technique
    that makes a copy of the part of the trail that was backtracked
    over. This saved copy can then be used to improve the efficiency
    of the solver's subsequent redescent. Although our technique does
    not save as much work as chronological backtracking, it also does
    not have chronological backtracking's shortcomings. Furthermore,
    the saved trail also provides the solver with the ability to look
    ahead along the previous trail. This can be exploited to improve
    its efficiency. Our empirical results show that our technique
    often yields superior performance over chronological backtracking,
    and that is able to improve the performance of state-of-the-art
    solvers.

\end{abstract}

\section{Introduction}
\textbf{Overall message of the paper.}

% \begin{enumerate}
% \item Formalism for understanding a more general form of trail savings
%     which caches a copy of the trail that is backtracked over, and
%     verifying how it exploited soundly.

% \item Potential advantages over chronological backtracking.  Literals
%     remain at the correct level, supports the solver having more
%     freedom in its decisions.

% \item Potential advantages of technique (1) increase in efficiency of
%     propagation, (2) the new uses of the cached information.

% \item Potential disadvantage. Reasons are recycled, in some circumstances
%     new reasons might be better. We show in a simple experiment that
%     this effect can occur.

% \item Technique captures most of the positive effects of
%     non-chronological backtracking in a simpler to implement way. It
%     also preserves the simpler standard structure of CDCL
%     solvers. Already provides some advantages for current SAT
%     solvers. The ideas also have the potential for further
%     exploitation. Lookahead for conflicts seems to be beneficial, and
%     potentially if better discrimatory measures could be developed for
%     reason clauses we could also better determine whether or not to
%     use the cached information.. 
% \end{enumerate}

% \note{Reusing the Assignment Trail in CDCL Solvers: restart to level
%   $i$ where the next decision has a higher score than the decision at
%   level $i$ and a lower score than all decisions at level $j$ for
%   $j< i$.

%   They argum changes to reason clauses don't make much difference, but
%   that was with minisat, before metrics for clause quality were in place in sat solvers.

%   The savings for Luby-100 (used in minisat) were about 1\%}

The vast majority of modern SAT solvers that are used to solve
real-world problems are based on the conflict-driven clause learning
(CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after
every conflict, where all literals from one or more decision levels
become unassigned, before the solver resumes making decisions and
performing unit propagations. Traditionally, CDCL solvers would
backtrack non-chronologically to the conflict level, which is the
second highest decision level remaining in the conflict clause after
conflict analysis has resolved away all but one literal from the
current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
however, it has been shown that partial backtracking
\cite{DBLP:conf/lpar/JiangZ13} or chronological backtracking, \cbt,
(i.e. backtracking one level only after conflict analysis)
\cite{DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19} can be effective
on many instances and was implemented in solvers that won the last two
SAT competitions. Although chronological backtracking breaks some of
the conventional invariants of CDCL solvers, it has been formalized
and proven correct \cite{DBLP:conf/sat/MohleB19}.

The motivation for using \cbt is the observation that when a solver
backtracks across many levels, many of the literals that are
unassigned during the backtrack will be re-assigned again in roughly
the same order when the solver redescends. This observation was first
made in the context of restarts by van der Tak et
al. \cite{DBLP:journals/jsat/TakRH11}. Their technique backtracks to
the minimum change level, i.e., the first level at which the solver's
trail can change on redescent. However, their technique cannot be used
when backtracking from a conflict: the solver's trail is going to be
changed at the backtrack level so the minimum change level is the same
as the backtrack level.

@@

In this paper we present an alternative ``trail saving'' method whereby
we save the backtracked part of the solver's trail and attempt to
use that information to make the solver's redescent more
efficient. Trail saving preserves the traditional invariants of the
SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and sometimes better than
chronological backtracking. We also show that with the enhancements we
are able to improve the performance of state-of-the-art solvers.

@@

\paragraph{Chronological Backtracking.}
Chronological bactracking (\cbt) is a technique originally proposed by
Jiang and Zhang \cite{DBLP:conf/lpar/JiangZ13} under the name
\textit{partial backtracking}. Its aim is to reduce the redundant work
that might be done by the \sat solver on its redescent from the
backtrack level $\btL$. The original paper proposed a technique for
backtracking to any level $j$ in the range
$\btL \leq j \leq \deepestLvl{-}1$ (where $\deepestLvl$ is the level
the conflict was discovered). Nadel and Ryvchin
\cite{DBLP:conf/sat/NadelR18} proposed to always backtrack to
$\deepestLvl{-}1$ while M{\"{o}}hle and Biere
\cite{DBLP:conf/sat/MohleB19} returned to the original proposal of
flexibly backtracking to any level in the allowed range. Note that the
new 1-UIP clause $\uipcls$ is made unit at every level in this
range. So after backtracking to level $j$ the newly implied literal
$\deepestLit$ can be added to the trail with
$\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$ is set to
$\btL$ (the second deepest level in $\uipcls$.

This means that the decision levels on the trail are no longer
contiguous, as $\deepestLit$ has a different level than the other
literals at level $j$ (if $j\neq \btL$). This change has a number of
consequences for the \sat solver's operation, all of which were
described in the original paper \cite{DBLP:conf/lpar/JiangZ13}. In
\cite{DBLP:conf/sat/NadelR18} some more practical implementation
methods were developed to overcome these issues, and in
\cite{DBLP:conf/sat/MohleB19} a formal framework for \cbt was
developed under which the changes required to the \sat solver could be
shown to be sound. We refer the reader to the cited papers for
more details.

\section{Potential Drawbacks of Chronological Backtracking}
In this paper we present a new technique that allows the \sat solver
to use standard backtracking, but also allows saving some redundant
work on its redescent.  Our method has more overhead than \cbt so the
first question that must be addressed is why not just use
chronological backtracking.

The reduction in redundant work achieved by \cbt does not come for
free. In particular, some of the work avoided by \cbt is not actually
redundant work; it is work that can actually help the solver. The
obvious evidence for this statement is that in both
\cite{DBLP:conf/sat/NadelR18} and \cite{DBLP:conf/sat/MohleB19} it was
found that it was not optimal to always apply \cbt. In
\cite{DBLP:conf/sat/NadelR18} \cbt was applied only when the length of
the standard backtrack, $\btL-\deepestLvl$ was greater than a given
threshold $T$. In their experiments they found that $T=100$ was the
best value, i.e., \cbt is done only on longer backtracks. In practice,
this meant that \cbt was \emph{hardly ever done}; in our measurements
with their solver only about 3\% of the solver backtracks were \cbt
backtracks. In \cite{DBLP:conf/sat/MohleB19} the value $T=100$ was
also applied. However, they introduced an additional technique to add
some applications of \cbt when the length of the backtrack is less
than $T$. This allowed \cite{DBLP:conf/sat/MohleB19} to utilize \cbt
in about \%15 of the backtracks. Nevertheless, the fact \cbt has to be used
in a controlled fashion, indicates that there can be benefits in the \sat
solver employing standard backtracking; it cannot be the case that
every redescent is only doing redundant work.

What is this non-redundant work? With standard backtracking the
literal $\deepestLit$ is placed on the trail at the end of $\btL$ and
then unit propagated. This could impact the trail in at least the
following ways. First, some literals might become unit at earlier
levels. This could include decision literals becoming forced which
might compress some decision levels together. Second, different
decisions might be made due to changes in the variable scores arising
from the newly learnt clause. And third, literals might be unit
implied with different reasons. These changes could have a major
impact on the future learnt clauses, and thus on the solver's
efficiency. Unfortunately, it seems difficult to predict when these
impacts might be large.

The second impact, changing variable scores, is partially addressed in
\cite{DBLP:conf/sat/MohleB19} who utilize the ideas of
\cite{DBLP:journals/jsat/TakRH11} to backtrack to a level where the
decisions would be unchanged. However, if the length of the backtrack
is greater than $100$ there could still be a divergence between the
variable decisions generated by standard backtracking and \cbt. An
argument is also given in \cite{DBLP:journals/jsat/TakRH11} that the
third impact, changing literal reasons, is not significant. However,
the experiments in \cite{DBLP:journals/jsat/TakRH11} were run before
good notions of clause quality were known
\cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results indicate
that once clause quality is accounted for changing the literal reasons
can have a significant impact.

The first impact is worth discussing since it was mentioned in the
original \cbt paper \cite{DBLP:conf/lpar/JiangZ13} but not in the
subsequent works. This is the issue of changing the decision levels of
literals on the trail. \cbt computes the decision level of each
implied literal based on the decision levels of its reasons, but it
does not go backwards to change the decisions levels of literals
earlier on the trail.

@@

In sum, although these impacts of \cbt might or might not be harmful
to the \sat solver, they exist. Hence, perhaps the most unappealing
aspect of \cbt is that it introduces additional interdependencies
between components of the solver. For example, the second impact means
that heuristics for choosing decision variables might work differently
in the presence of \cbt; and the first impact means that different
ways of computing clause quality using the decision levels in the
clauses might also work differently in the presence of \cbt. So
although there are bound to be cases where \cbt is superior, we would
like to have a method that addresses the original motivation of
reducing redundant work while not changing the operation of standard
backtracking. We present our proposal for such a method next.

@@

In Figure~\ref{fig:cadicalNew} we used the newest version of cadical
(downloaded as of January 1, 2020) as the baseline solver, in
Figure~\ref{fig:cadicalOld} we used the version of cadical published
in \cite{DBLP:conf/sat/MohleB19} as the baseline, and in
Figure~\ref{fig:maple} we used MapleLCMDist \cite{xiao2017maplelrb} as
the baseline. Each of the baselines were run with standard
non-chronological backtracking. We then refer to versions of each
solver with additional features implemented on top by adding
suffixes. "-chrono" refers to the solver with \cbt enabled (using the
solver's default settings), "-trail" refers to the baseline with plain
trail saving added (as described in Figure~\ref{fig:trailUse}),
"-trail-multipleBT" refers to the baseline with trail saving plus the
first enhancement of saving over multiple backtracks,
"-trail-multipleBT-lookahead" also adds the enhancement of lookahead
for conflicts by 2 decision levels, and
"-trail-multipleBT-lookahead-reason" also adds the final enhancement
to cease trail saving once a reason of "low quality" is reached. For
more details on the enhancements, please see section 4.2.

>>>>

\begin{document}
\maketitle
\begin{abstract}
    A CDCL \sat solver can backtrack a large distance when it learns a
    new clause, e.g, when the new learnt clause is a unit clause the
    solver has to backtrack to level zero. When the length of the
    backtrack is large, the solver can end up reproducing many of the
    same decisions and propagations when it redescends the search
    tree. Different techniques have been proposed to reduce this
    potential redundancy, e.g., partial/chronological backtracking and
    trail saving on restarts. In this paper we present a new trail
    saving technique that is not restricted to restarts, unlike prior
    trail saving methods. Our technique makes a copy of the part of
    the trail that is backtracked over. This saved copy can then be
    used to improve the efficiency of the solver's subsequent
    redescent. Furthermore, the saved trail also provides the solver
    with the ability to look ahead along the previous trail which can
    be exploited to improve its efficiency. Our new trail saving
    technique offers different tradeoffs in comparison with
    chronological backtracking and often yields superior performance.
    We also show that our technique is able to improve the performance
    of state-of-the-art solvers.
\end{abstract}

\section{Introduction}
The vast majority of modern SAT solvers that are used to solve
real-world problems are based on the conflict-driven clause learning
(CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after
every conflict, where all literals from one or more decision levels
become unassigned before the solver resumes making decisions and
performing unit propagations. Traditionally, CDCL solvers would
backtrack to the conflict level, which is the second highest decision
level remaining in the conflict clause after conflict analysis has
resolved away all but one literal from the current decision level
\cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently, however, it has been
shown that partial backtracking \cite{DBLP:conf/lpar/JiangZ13} or
chronological backtracking, \cbt, (i.e., backtracking only to the
previous level after conflict analysis)
\cite{DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19} can be effective
on many instances. Partial backtracking has been used in the solvers
that won the last two SAT competitions. Although chronological
backtracking breaks some of the conventional invariants of CDCL
solvers, it has been formalized and proven correct
\cite{DBLP:conf/sat/MohleB19} (also see related formalizations 
\cite{DBLP:journals/jacm/NieuwenhuisOT06,nadelPhD}).

The motivation for using \cbt is the observation that when a solver
backtracks across many levels, many of the literals that are
unassigned during the backtrack might be re-assigned again in roughly
the same order when the solver redescends. This observation was first
made in the context of restarts by van der Tak et
al. \cite{DBLP:journals/jsat/TakRH11}. Their technique backtracks to
the minimum change level, i.e., the first level at which the solver's
trail can change on redescent. However, their technique cannot be used
when backtracking from a conflict: the solver's trail is going to be
changed at the backtrack level so the minimum change level is the same
as the backtrack level.

@@

In this paper we present an new trail saving method whereby we save
the backtracked part of the solver's trail and attempt to use that
information to make the solver's redescent more efficient. Unlike
\cbt, our trail saving method preserves the traditional invariants of
the SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and often better than
chronological backtracking. We also show that with our enhancements we
are able to improve the performance of state-of-the-art solvers.

@@

\paragraph{Chronological Backtracking.}
Chronological backtracking (\cbt) and partial backtracking in the
context of clause learning solvers are alternatives to standard
backtracking which allow the solver to execute a shorter
backtrack. That is, with these techniques the solver can avoid having
to go all the way back to the second deepest level in the learnt
clause, as in standard backtracking.

Formalisms for partial backtracking in clause learning solvers have
been presented in
\cite{DBLP:journals/jacm/NieuwenhuisOT06,nadelPhD}. In
\cite{DBLP:conf/lpar/JiangZ13} practical issues of implementation were
addressed, and experiments shown with a CDCL solver using partial
backtracking. In \cite{DBLP:conf/sat/NadelR18} improved and more
efficient implementation techniques were developed which allowed \cbt
to make improvements to state-of-the-art \sat solvers, and
\cite{DBLP:conf/sat/MohleB19} presented additional implementation
ideas and details along with correctness results for these methods.

The aim of partial backtracking is to reduce the redundant work that
might be done by the \sat solver on its redescent from the backtrack
level $\btL$. The technique allows the solver to backtrack to any
level $j$ in the range $\btL \leq j \leq \deepestLvl{-}1$ (where
$\deepestLvl$ is the level the conflict was discovered). Nadel and
Ryvchin \cite{DBLP:conf/sat/NadelR18} proposed to always backtrack
chronologically to $\deepestLvl{-}1$ while M{\"{o}}hle and Biere
\cite{DBLP:conf/sat/MohleB19} returned to the proposal of
\cite{DBLP:conf/lpar/JiangZ13} of flexibly backtracking to any level
in the allowed range. Note that the new learnt 1-UIP clause $\uipcls$
is made unit at every level in this range. So after backtracking to
level $j$ the newly implied literal $\deepestLit$ is added to the
trail with $\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$
is set to $\btL$ (the second deepest level in $\uipcls$).

This means that the decision levels on the trail are no longer
contiguous, as $\deepestLit$ has a different level than the other
literals at level $j$ (if $j\neq \btL$). This change has a number of
consequences for the \sat solver's operation, all of which were
described in \cite{DBLP:conf/lpar/JiangZ13}. M{\"{o}}hle and Biere
\cite{DBLP:conf/satt/MohleB19} showed that despite these consequences
partial backtracking can be made to preserve the soundness of a CDCL
solver.

\section{Chronological Backtracking Effects on Search}
In this paper we present a new technique that allows the \sat solver
to use standard backtracking, but also allows saving some redundant
work on its redescent.  Our method has more overhead than \cbt so the
first question that must be addressed is why not just use
chronological backtracking.

Although \cbt is able to avoid a lot of redundant work it also has
other effects on the \sat solver search. These effects are sometimes
detrimental to the solver's performance and so it is not always
beneficial to use \cbt. In fact, in both \cite{DBLP:conf/sat/NadelR18}
and \cite{DBLP:conf/sat/MohleB19} it was found that fairly limited
application of \cbt performed best. In \cite{DBLP:conf/sat/NadelR18}
\cbt was applied only when the length of the standard backtrack,
$\btL-\deepestLvl$ was greater than a given threshold $T$. In their
experiments they found that $T=100$ was the best value, i.e., \cbt is
done only on longer backtracks. In practice, this meant that \cbt was
relatively infrequent; in our measurements with their solver only
about 3\% of the solver backtracks were \cbt backtracks. In
\cite{DBLP:conf/sat/MohleB19} the value $T=100$ was also
applied. However, they introduced an additional technique to add some
applications of \cbt when the length of the backtrack is less than
$T$. This allowed \cite{DBLP:conf/sat/MohleB19} to utilize \cbt in
about 15\% of the backtracks.

Although it is difficult to know precisely why \cbt is not always
beneficial, we can identify some different ways in which \cbt can
affect the \sat solver's search. With standard backtracking the
literal $\deepestLit$ is placed on the trail at the end of $\btL$ and
then unit propagated. This could impact the trail in at least the
following ways. First, some literals might become unit at earlier
levels. This could include decision literals becoming forced which
might compress some decision levels together. Second, different
decisions might be made due to changes in the variable scores arising
from the newly learnt clause. And third, literals might be unit
implied with different reasons. \cbt can change all of these things,
each of which could have an impact on the future learnt clauses, and
thus on the solver's overall efficiency.

The second impact, changing variable scores, is partially addressed in
\cite{DBLP:conf/sat/MohleB19} who utilize the ideas of
\cite{DBLP:journals/jsat/TakRH11} to backtrack to a level where the
decisions would be unchanged. However, if the length of the backtrack
is greater than $100$ there could still be a divergence between the
variable decisions generated in standard backtracking and \cbt. An
argument is also given in \cite{DBLP:journals/jsat/TakRH11} that the
third impact, changing literal reasons, is not significant. However,
the experiments in \cite{DBLP:journals/jsat/TakRH11} were run before
good notions of clause quality were known
\cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results indicate
that once clause quality is accounted, for changing the literal
reasons can have a significant impact.

The first impact is worth discussing since it was mentioned in
\cite{DBLP:conf/lpar/JiangZ13} but not in the subsequent works. This
is the issue of changing the decision levels of literals on the
trail. \cbt computes the decision level of each implied literal based
on the decision levels of the literals in its reason, but it does not
go backwards to change the decisions levels of literals earlier on the
trail.

@@

In sum, although these impacts of \cbt on the \sat solver's search
might or might not be harmful to the \sat solver, they do exist. In
fact, there are two pieces of evidence that these impacts can
sometimes be harmful. First, as mentioned above, previous work found
that it is best to only apply \cbt on large backtracks where it has
the potential to save the most work. If there were no harmful effects
it would always be effective to apply \cbt. And second, in our
empirical results below we show that our new trail saving technique,
which always uses standard backtracking, can often outperform
\cbt. Although our technique reduces the solver's work on redescent it
does not completely eliminate it like \cbt does. Hence its superior
performance can only occur if \cbt is sometimes harmful.

It is possible to combine \cbt with our trail saving technique to
reduce the amount of work required whenever the solver performs
non-chronological backtracking. However, \cbt greatly reduces the
potential savings that could be achieved by our method since most of
its non-chronological backtracks are relatively short (less than
threshold $T$ levels). In our preliminary experiments this combination
did not seem promising.

Nevertheless, there is good evidence that \cbt can improve \sat solver
performance.\footnote{\cbt can also be extremely useful in contexts
  where each descent can be very expensive, e.g., when doing theory
  propagation in SMT solving, or component analysis in \#\sat
  solving. In these cases, \cbt, by avoiding backtracking and
  subsequent redescent, has considerable potential for improving
  solver performance.} Hence, it should be that it is better to
perform \cbt in some branches. Hence, an interesting direction for
future work would be to develop better heuristics about when to use
\cbt in a branch and when to use standard backtracking augmented by
our trail saving method.

@@

In Figure~\ref{fig:cadicalNew} we used the newest version of cadical
(downloaded as of January 1, 2020) as the baseline solver, in
Figure~\ref{fig:cadicalOld} we used the version of cadical published
in \cite{DBLP:conf/sat/MohleB19} as the baseline, and in
Figure~\ref{fig:maple} we used MapleLCMDist \cite{DBLP:conf/ijcai/LuoLXML17,xiao2017maplelrb} as
the baseline. Each of the baselines were run with standard
non-chronological backtracking. We then refer to versions of each
solver with additional features implemented on top by adding
suffixes. "-chrono" refers to the solver with \cbt enabled (using the
solver's default settings), "-trail" refers to the baseline with plain
trail saving added (as described in Figure~\ref{fig:trailUse}),
"-trail-multipleBT" refers to the baseline with trail saving plus the
first enhancement of saving over multiple backtracks,
"-trail-multipleBT-lookahead" also adds the enhancement of lookahead
for conflicts by 2 decision levels, and
"-trail-multipleBT-lookahead-reason" also adds the final enhancement
to cease trail saving once a reason of "low quality" is reached. For
more details on the enhancements, please see section 4.2.
